Predicting Water Treatment Plant Chemical Doses

### 2025-12-11 Data Cleaning.Rmd
Updates to classification methods

# Model fit to take performance metrics of:
rf_fit <- rf_fit3

@@

## Accuracy
rf_fit |>
  augment(new_data = df_test) |>
  accuracy(truth = FeCl_change, estimate = .pred_class)
## Test confusion matrix
rf_fit |>
  augment(new_data = df_test) |>
  conf_mat(truth = FeCl_change, estimate = .pred_class)

## Notes
# 34% accuracy for all predictors
# 46.6% accuracy for just temperature and pH

>>>>

# Model fit to take performance metrics of:
rf_fit <- rf_fit2

@@

## Accuracy
rf_acc <- rf_fit |>
  augment(new_data = df_test) |>
  accuracy(truth = FeCl_change, estimate = .pred_class)
## Test confusion matrix
rf_conf <- rf_fit |>
  augment(new_data = df_test) |>
  conf_mat(truth = FeCl_change, estimate = .pred_class)

rf_acc
rf_conf

## Notes
# 34% accuracy for all predictors
# 39% accuracy for temp + pH + temp/pH lags
# 46.6% accuracy for just temperature and pH

### 2025-12-11 Presentation DSCI445.Rmd
Updates to classification methods

```{r, include=F}
library(dplyr)
library(tidyr)
library(readr)
library(lubridate)
library(ggplot2)
library(zoo)
library(purrr)
library(knitr)
library(patchwork)
library(tidyverse)
library(tidymodels)
library(rpart.plot)
library(kableExtra)
library(vip)
library(png)
library(grid)
```

@@

The drinking water treatment process takes water from a river, lake, reservoir, or other source, and purifies it to have it reach drinking water standards.

@@

```{r, fig.width=7, fig.height=2}
df_lag_incdec <- read.csv("RWAR_dosing_lag_incdec.csv") |>
  mutate(Timestamp = ymd(Timestamp))

@@

# Summary of Models

@@

- Training and Testing
   - Training: 2018-2019
   - Testing: 2020
   - Approximately 70/30 split
- Coagulant Dose Prediction
   - Linear Regression
   - Tree-based Models
   - GAM
- Coagulant Dose Change Prediction
   - Multinomial Classification
   - LDA, QDA
   - Random Forest

@@

- Performed best with multiple interactions between Conductivity, Temp, and pH

@@

- Linear Structure doesn't capture the relations in our predictors, and incorrect predictions could lead to disastorus results if too much or too little chemical is used.

@@

# Coagulant Dose Change Prediction: Multinomial, LDA/QDA

@@

>>>>

```{r, include=F}
library(dplyr)
library(tidyr)
library(readr)
library(lubridate)
library(ggplot2)
library(zoo)
library(purrr)
library(knitr)
library(patchwork)
library(tidyverse)
library(tidymodels)
library(rpart.plot)
library(kableExtra)
library(vip)
library(png)
library(grid)
library(parsnip)
library(discrim)
```

```{r}
df_lag_incdec <- read.csv("RWAR_dosing_lag_incdec.csv") |>
  mutate(Timestamp = ymd(Timestamp))
df_lag_incdec$FeCl_change <- as.factor(df_lag_incdec$FeCl_change)
df_lag_incdec$Polymer_change <- as.factor(df_lag_incdec$Polymer_change)
```

@@

**The drinking water treatment process takes water from a river, lake, reservoir, or other source, and purifies it to have it reach drinking water standards.**

@@

```{r, fig.width=7, fig.height=2}

@@

# Summary of Modeling

@@

- Training and Testing
   - Training: 2018-2019
   - Testing: 2020
   - Approximately 70/30 split
- Coagulant Dose Prediction
   - Linear Regression
   - Tree-based Models
   - GAM
- Coagulant Dose Change Prediction
   - Linear methods: Multinomial, LDA
   - Nonlinear methods: QDA, Random Forest

@@

- Performed best with multiple interactions between Conductivity, Temp, and pH.

@@

- Linear structure doesn't capture the relations in our predictors.

@@

# Coagulant Dose Change Prediction: LDA

```{r}
set.seed(445)

## TRAIN/TEST DATA (2018-2019 and 2020)
df_full <- df_lag_incdec |> select(-c(FeCl_dose, Polymer_dose))
df_train <- df_full |> filter(year(Timestamp) <= 2019)
df_test  <- df_full |> filter(year(Timestamp) == 2020)

## LDA
lda_spec <- discrim_linear()
lda_fit1 <- lda_spec |>
  fit(FeCl_change ~ ., data = df_train |> 
        select(-c(Polymer_change, Timestamp)))
lda_fit2 <- lda_spec |>
  fit(FeCl_change ~ mean_pH + mean_Temp + pH_lag1 + pH_lag2 + Temp_lag1 + Temp_lag2, data = df_train)

# select LDA to take metrics of
lda_fit <- lda_fit2

# performance metrics
lda_fit_disp <- lda_fit |> extract_fit_engine()
lda_conf <- lda_fit |> 
  augment(new_data = df_test) |>
  conf_mat(truth = FeCl_change, estimate = .pred_class)
lda_acc <- lda_fit |>
  augment(new_data = df_test) |>
  accuracy(truth = FeCl_change, estimate = .pred_class)

lda_table <- lda_conf$table |> 
  as.data.frame() |>
  rename(Predicted = Prediction, True = Truth, Count = Freq)
lda_table |>
  tidyr::pivot_wider(names_from = Predicted, values_from = Count) |>
  knitr::kable("markdown", caption = "LDA Confusion Matrix")

## Notes
# 39% test accuracy on all predictors
# 41% test accuracy on temp and pH

# Coagulant Dose Change Prediction: QDA

```{r}
set.seed(445)

## QDA
qda_spec <- discrim_quad()
qda_fit1 <- qda_spec |>
  fit(FeCl_change ~ ., data = df_train |> 
        select(-c(Polymer_change, Timestamp)))
qda_fit2 <- qda_spec |>
  fit(FeCl_change ~ mean_pH + mean_Temp + pH_lag1 + pH_lag2 + Temp_lag1 + Temp_lag2, data = df_train)

# select LDA to take metrics of
qda_fit <- qda_fit2

# performance metrics
qda_fit_disp <- qda_fit |> extract_fit_engine()
qda_conf <- qda_fit |> 
  augment(new_data = df_test) |>
  conf_mat(truth = FeCl_change, estimate = .pred_class)
qda_acc <- qda_fit |>
  augment(new_data = df_test) |>
  accuracy(truth = FeCl_change, estimate = .pred_class)

qda_table <- qda_conf$table |> 
  as.data.frame() |>
  rename(Predicted = Prediction, True = Truth, Count = Freq)
qda_table |>
  tidyr::pivot_wider(names_from = Predicted, values_from = Count) |>
  knitr::kable("markdown", caption = "QDA Confusion Matrix")

## Notes
# 39.5% test accuracy on all predictors
# 43.7% test accuracy on temp and pH

# Coagulant Dose Change Prediction: Multinomial Classification

```{r}
## MULTINOMIAL REGRESSION
multi_model <- multinom_reg() |>
  set_engine("nnet") |>
  set_mode("classification")

multi_recipe1 <- recipe(FeCl_change ~ mean_Temp + Temp_lag1 + Temp_lag2 + mean_pH + pH_lag1 + pH_lag2 + mean_Conductivity + Cond_lag1 + Cond_lag2, data = df_train) |>
  step_normalize(all_numeric_predictors()) |>
  step_dummy(all_nominal_predictors())

multi_recipe <- multi_recipe1

multi_workflow <- workflow() |>
  add_recipe(multi_recipe) |>
  add_model(multi_model)

multi_fit <- fit(multi_workflow, data = df_train)

df_pred_class <- predict(multi_fit, new_data = df_test, type = "class") |>
  bind_cols(df_test |> select(FeCl_change))
df_pred_probs <- predict(multi_fit, new_data = df_test, type = "prob") |>
  bind_cols(df_test |> select(FeCl_change))

# results
multi_acc <- metrics(df_pred_class, truth = FeCl_change, estimate = .pred_class) |> 
  filter(.metric == "accuracy")
multi_conf <- conf_mat(df_pred_class, truth = FeCl_change, estimate = .pred_class)
multi_fit_disp <- multi_fit |> extract_fit_engine()

multi_table <- multi_conf$table |> 
  as.data.frame() |>
  rename(Predicted = Prediction, True = Truth, Count = Freq)
multi_table |>
  tidyr::pivot_wider(names_from = Predicted, values_from = Count) |>
  knitr::kable("markdown", caption = "Multinomial Regression Confusion Matrix")

## NOTES
# turbidity is highly variable and does not help

@@

```{r}

set.seed(445)

## RANDOM FOREST
rf_spec <- rand_forest(mtry = sqrt(.cols()), trees=500, min_n=1) |>
  set_engine("randomForest", importance = TRUE) |>
  set_mode("classification")

# Different model fits
rf_fit1 <- rf_spec |>
  fit(FeCl_change ~ ., data = df_train |> 
        select(-c(Polymer_change, Timestamp)))
rf_fit2 <- rf_spec |>
  fit(FeCl_change ~ mean_Temp + mean_pH + Temp_lag1 + Temp_lag2 + pH_lag1 + pH_lag2, 
      data = df_train)
rf_fit3 <- rf_spec |>
  fit(FeCl_change ~ mean_pH + mean_Temp, 
      data = df_train)

# Model fit to take performance metrics of:
rf_fit <- rf_fit2

# Most important predictors
#vip(rf_fit)

## results
rf_acc <- rf_fit |>
  augment(new_data = df_test) |>
  accuracy(truth = FeCl_change, estimate = .pred_class)
rf_conf <- rf_fit |>
  augment(new_data = df_test) |>
  conf_mat(truth = FeCl_change, estimate = .pred_class)

rf_table <- rf_conf$table |> 
  as.data.frame() |>
  rename(Predicted = Prediction, True = Truth, Count = Freq)
rf_table |>
  tidyr::pivot_wider(names_from = Predicted, values_from = Count) |>
  knitr::kable("markdown", caption = "Random Forest Confusion Matrix")

## Notes
# 34% accuracy for all predictors
# 39% accuracy for temp + pH + temp/pH lags
# 46.6% accuracy for just temperature and pH
